Numerical simulation of bubble deformation \\ and breakup under simple linear shear flows

### 2022-12-26 Bubble-deformation.tex
incorporated Fig12 from Mitsuhiro, added mathematical description of the ICs and BCs

Figure~\ref{fig:SchemAndGrid}(a) shows a schematic of the computational system
for our studies of a bubble (or drop) in shear flow.  The computational domain
consists of a three-dimensional rectangular domain of $L(x)$ (length) $\times$
$W(y)$ (width) $\times$ $H(z)$ (height).  The size of $L(x)$, $W(y)$ and $H(z)$
was determined after consideration of numerical result sensitivity to domain
size; numerical studies of domain-size dependence are presented in
Section~\ref{sec:DomGrdSize}.  All computational results that follow were
obtained from numerical solutions of the three-dimensional governing equations
for gas-liquid/liquid-liquid flows.  Computations are initialized with a
spherical bubble set at the center of the domain that is then subjected to a
linear shear flow generated by the motion of the top and bottom plates, which 
have constant velocity $+V$ and $-V$, respectively.  In the interior of the
domain, the initial velocity condition is assumed to be a simple linear profile
and periodic boundary conditions are imposed along the $x$ and $y$ directions.
Common dimensionless physical parameters used to describe gas-liquid or
liquid-liquid two-phase flows include the Reynolds ($Re$), Weber ($We$) (or
$Ca$ (= $We/Re$)) and Froude ($Fr$) numbers, and the density ratio $\lambda$
and viscosity ratio $\eta$.  For comparison with previous drop studies, we set
$\lambda=1$ and, since in this case the effect of gravity is not considered, we
may ignore the Froude number $Fr$.  As a result, (for $\lambda=1$) the
following dimensionless physical parameters are used to describe the problem of
drop or bubble deformation/breakup in shear flow
%
\begin{eqnarray}\label{dimensonless}
  Re = \frac{\rho_{b}UR}{\mu_{b}}, \quad
  Ca = \frac{\mu_{b}U}{\sigma}, \quad
  \eta = \frac{\mu_{b}}{\mu_{m}}
\end{eqnarray} 
%
where $R$ is the drop or bubble radius, $\sigma$ denotes the surface tension,
and $U$ is the velocity scale. For the problem of shear-induced drop
deformation and breakup, the velocity is set to $U = \mathit{\Gamma} R$, where
the shear-rate $\mathit{\Gamma} = 2V/H$.  As mentioned in the introduction,
most previous drop studies set $\eta = 1$. Thus, for comparison, in this case
we will set $\lambda = \eta = 1$ (and also neglect the effect of gravity so
that $g=0$). On the other hand, in our computations for a bubble, we set the
density and viscosity of air to $\rho_{b}$ = 1.2 kg/m$^{3}$ and $\mu_{b}$ = 1.8
$\times$ $10^{-5}$ Pa$\cdot$s, and we take the 
density ratio 
%%$\lambda$ = 1.2 $\times$ $10^{-3}$ 
$\lambda$ = 1.0 $\times$ $10^{-3}$ 
and the 
viscosity ratio 
%%$\eta$ $<$ 1.0 $\times$ $10^{-3}$.
$\eta$ $=$ 1.0 $\times$ $10^{-3}$.
For consistency with previous studies, we will computationally examine the
deformation and breakup of a bubble in simple linear shear flow as a
function of $Re$ and $Ca$ numbers.  In particular, 
for a certain $Ca$ value, we
focus on determining the critical value 
$Re_{c}$ that corresponds to the threshold
for breakup.

@@

In Figure~\ref{fig:CaRecFit} we plot our best fit curve and make the
hypothesis that a given $(Ca,Re)$ pair will result in bubble break up
if the point $(Ca,Re)$ is above the critical curve.

@@

%% variance for X^T X c = b
%% (S/(n-m)) (X^T X)^{-1}
\begin{figure}%[h!]
  \centering
  \includegraphics[width=\textwidth]{critical_Reynolds_Number_label}
  \caption{A plot of the critical Reynolds number versus Capillary number
	for the data listed in Table \ref{tab:CaRecComparison}.  The
	linear least squares best fit for the $\log(Re_{c})$ is
	$\log(Re_{c})=4.9-1.41(Ca)$ (i.e. $Re_{c}=138.3 e^{-1.41 Ca}$)
	with a predicted variance
	for the parameters ``$4.9$'' and ``$-1.41$'' being
	$4.2E-3$ and $8.4E-3$ respectively.  Even for the
	given data that we computed, it is suggested that 
	a ``bubble break-up map''
	can be derived, in which for the conditions that
	$\lambda$ = $\eta$ = 1.0 $\times$ $10^{-3}$ and $0<Ca<1$, one
	can predict whether a given $(Ca,Re)$ pair will result in shear
	induced bubble break-up or not.  
	   }
  \label{fig:CaRecFit}
\end{figure}
%

>>>>

Figure~\ref{fig:SchemAndGrid}(a) shows a schematic of the computational system
for our studies of a bubble (or drop) in shear flow.  The computational domain
consists of a three-dimensional rectangular domain of $L(x)$ (length) $\times$
$W(y)$ (width) $\times$ $H(z)$ (height).  The size of $L(x)$, $W(y)$ and $H(z)$
was determined after consideration of numerical result sensitivity to domain
size; numerical studies of domain-size dependence are presented in
Section~\ref{sec:DomGrdSize}.  All computational results that follow were
obtained from numerical solutions of the three-dimensional governing equations
for gas-liquid/liquid-liquid flows.  Computations are initialized with a
spherical bubble set at the center of the domain that is then subjected to a
linear shear flow generated by the motion of the top and bottom plates, which 
have constant velocity $+V$ and $-V$, respectively.  In the interior of the
domain, the initial velocity condition is assumed to be a simple linear profile
and periodic boundary conditions are imposed along the $x$ and $y$ directions.
Mathematically, the initial and boundary conditions are described as
follows:
\begin{eqnarray}
	\phi(x,y,z,0)=\sqrt{(x-\frac{L}{2})^{2}+(y-\frac{W}{2})^{2}+
	(z-\frac{H}{2})^{2}}-R \label{IC_BC} \\
	\bmu(x,y,z,0)=\left( \begin{array}{c}
		\frac{2V}{H}(z-\frac{H}{2}) \\ 0 \\ 0 
	\end{array} \right) \nonumber \\
	\phi(x+L,y,z,t)=\phi(x,y,z,t) \hspace{20pt}
	\phi(x,y+W,z,t)=\phi(x,y,z,t) \nonumber \\
	\bmu(x,y,H,t)=\left( \begin{array}{c}
		                V \\ 0 \\ 0 
	\end{array} \right)  \hspace{20pt}
	\bmu(x,y,0,t)=\left( \begin{array}{c}
		                -V \\ 0 \\ 0 
	\end{array} \right)  \nonumber \\
	\bmu(x+L,y,z,t)=\bmu(x,y,z,t) \hspace{20pt} 
	\bmu(x,y+W,z,t)=\bmu(x,y,z,t). \nonumber
\end{eqnarray}
$\phi(x,y,z,t)$ is a material indicator function which is positive 
in the ``matrix'' fluid and negative in the bubble (or drop) fluid.
$\bmu(x,y,z,t)$ is the velocity.

Common dimensionless physical parameters used to describe gas-liquid or
liquid-liquid two-phase flows include the Reynolds ($Re$), Weber ($We$) (or
$Ca$ (= $We/Re$)) and Froude ($Fr$) numbers, and the density ratio $\lambda$
and viscosity ratio $\eta$.  For comparison with previous drop studies, we set
$\lambda=1$ and, since in this case the effect of gravity is not considered, we
may ignore the Froude number $Fr$.  As a result, (for $\lambda=1$) the
following dimensionless physical parameters are used to describe the problem of
drop or bubble deformation/breakup in shear flow
%
\begin{eqnarray}\label{dimensonless}
  Re = \frac{\rho_{b}UR}{\mu_{b}}, \quad
  Ca = \frac{\mu_{b}U}{\sigma}, \quad
  \eta = \frac{\mu_{b}}{\mu_{m}}
\end{eqnarray} 
%
where $R$ is the drop or bubble radius, $\sigma$ denotes the surface tension,
and $U$ is the velocity scale. For the problem of shear-induced drop
deformation and breakup, the velocity is set to $U = \mathit{\Gamma} R$, where
the shear-rate $\mathit{\Gamma} = 2V/H$.  As mentioned in the introduction,
most previous drop studies set $\eta = 1$. Thus, for comparison, in this case
we will set $\lambda = \eta = 1$ (and also neglect the effect of gravity so
that $g=0$). On the other hand, in our computations for a bubble, we set the
density and viscosity of air to $\rho_{b}$ = 1.2 kg/m$^{3}$ and $\mu_{b}$ = 1.8
$\times$ $10^{-5}$ Pa$\cdot$s, and we take the 
density ratio 
%%$\lambda$ = 1.2 $\times$ $10^{-3}$ 
$\lambda$ = 1.0 $\times$ $10^{-3}$ 
and the 
viscosity ratio 
%%$\eta$ $<$ 1.0 $\times$ $10^{-3}$.
$\eta$ $=$ 1.0 $\times$ $10^{-3}$.
For consistency with previous studies, we will computationally examine the
deformation and breakup of a bubble in simple linear shear flow as a
function of $Re$ and $Ca$ numbers.  In particular, 
for a certain $Ca$ value, we
focus on determining the critical value 
$Re_{c}$ that corresponds to the threshold
for breakup.

@@

In Figure~\ref{fig:CaRecFit} we plot our best fit curve and make the
hypothesis that a given $(Ca,Re)$ pair will result in bubble break up
if the point $(Ca,Re)$ is above the critical curve, 
$0.3<Ca<1$, and the initial/boundary conditions are given
by (\ref{IC_BC}).

@@

%% variance for X^T X c = b
%% (S/(n-m)) (X^T X)^{-1}
\begin{figure}%[h!]
  \centering
  \includegraphics[width=\textwidth]{Fig12-Critical-Re}
  \caption{A plot of the critical Reynolds number versus Capillary number
	for the data listed in Table \ref{tab:CaRecComparison}.  The
	linear least squares best fit for the $\log(Re_{c})$ is
	$\log(Re_{c})=4.9-1.41(Ca)$ (i.e. $Re_{c}=138.3 e^{-1.41 Ca}$)
	with a predicted variance
	for the parameters ``$4.9$'' and ``$-1.41$'' being
	$4.2E-3$ and $8.4E-3$ respectively.  Even for the
	given data that we computed, it is suggested that 
	a ``bubble break-up map''
	can be derived, in which for the conditions that
	$\lambda$ = $\eta$ = 1.0 $\times$ $10^{-3}$, $0.3<Ca<1$, and
        the initial/boundary conditions are given by (\ref{IC_BC}),	
	one
	can predict whether a given $(Ca,Re)$ pair will result in shear
	induced bubble break-up or not.  
	   }
  \label{fig:CaRecFit}
\end{figure}
%
